Boundary terms in the action for Causal Sets

### 2014-10-16 paper_v7.tex
changes to start of chapter two and to rnc part

In order to give intuition to the ideas laid out in this paper it will be most convenient to consider spacetimes with boundary obtained by partitioning a manifold (without boundary) into two parts as follows. Let $(M,g)$ be a sufficiently well-behaved $d$-dimensional Lorentzian manifold. We assume that $(M,g)$ has finite volume and admits a \emph{spacelike partitioning} surface, $\Sigma$, defined as a spacelike closed compact surface in $M$, such that the causal past and future sets, $M^\pm=J^\pm (\Sigma)$ satisfy $M^+\cap M^-=\Sigma$. It then follows that $\partial M^\pm = \Sigma$, and further that no points in $M^+\setminus\Sigma$ precede any points in $M^-\setminus\Sigma$.~\!\footnote{If there did exist a $p\in M^+\setminus\Sigma$ and $q\in M^-\setminus\Sigma$ such that $p\preceq q$, then we would have $(q\in J^-(\Sigma) \text{ and } p\preceq q)\implies p\in J^-(\Sigma)\implies p\in M^-$, which would contradict $M^+\cap M^-=\Sigma$.} The Gibbons-Hawking-York boundary term for $M^\pm$ is then given by
\be\label{eq:GHYBT_in_continuum}
{S}^{(d)}_{GHY}\left[M^\pm\right]= \mp \frac{1}{l_p^{d-2}}\int_{\Sigma} K d\Sigma
\ee
where $l_p$ is the rationalised Planck length $l_p=(8\pi G)^{\frac{1}{d-2}}$ ($G$ denoting the gravitational constant in $d$ dimensions) and where $K$ is the trace of the extrinsic curvature $K_{\mu\nu}=h_{\mu}^\rho h_\nu^\sigma \nabla_\rho n_\sigma$ of $\Sigma$, defined here with future-pointing timelike unit normal covector $n_{\mu}=\partial_\mu S/\sqrt{g^{\mu\nu}\partial_\mu S\partial_\nu S}$. (We will work with a mostly plus convention for the metric, in which this translates into a past-pointing normal vector $n^{\mu}$).

@@

Fix $p\in M$ and denote its coordinate values in GNCs by $x^\mu_p= (t_p,\mathbf x_p)$. It will be convenient to use RNCs centered not at the tip, $p$, of the cone but instead at the point $p_0$ where the unique geodesic through $p$ whose tangent is normal to $\Sigma$ intersects $\Sigma$. In GNCs this simply corresponds to the point $p_0$ with coordinates $x_0^\mu=x^\mu (p_0)= (0,\mathbf x_p)$. RNCs centered at $p_0$ will be given the symbol $y^{\overline{\mu}}$, or in terms of coordinates as $(y^{\overline{0}},\mathbf{y})=(\overline{t},\mathbf{y})$.
%\mb{let's put all facts about $x^\mu$ and $y^{\overline\mu}$ for $p$ and $p_0$ here and expansion of det in RNC.} 
We need to assume that for any $p$ in $U_\varepsilon$, the Riemann normal neighbourhood $U_p\subset M$ (throughout which the RNC system centered at $p_0$ is well-defined) contains the truncated cone $J^- (p)\cap J^+ (\Sigma)$. This assumption seems reasonable given the that $U_\varepsilon$ can be made arbitrarily ``thin'' as $\rho\rightarrow\infty$. In RNCs the metric and the Christoffel symbols at $p_0$ are those of flat space:
\be\label{eq:RNCMetricTransAtPAndChris}
g_{\overline{\mu} \overline{\nu}} (p_0)=\eta_{\overline{\mu} \overline{\nu}}=A^{\mu}_{\;\overline{\mu}}A^{\nu}_{\;\overline{\nu}}g_{\mu\nu} (p_0)\;,\;\;\;\;\Gamma^{\overline{\mu}}_{\;\overline{\nu}\overline{\rho}} (p_0)=0
\ee
The $A^{\mu}_{\;\overline{\mu}}$ govern the coordinate transformation from GNCs to RNCs to linear order. %but $O (x^2)$ corrections may be required. 
To second order the coordinate transformation is given by
\be\label{eq:RNCtotaltrans}
y^{\overline{\mu}}=A^{\overline{\mu}}_{\;\nu}(x^\nu -x_0^{\nu})+\frac{1}{2}A^{\overline{\mu}}_{\;\mu}\Gamma^{\mu}_{\;\nu\rho} (p_0)(x^\nu -x_0^{\nu}) (x^\rho -x_0^{\rho})+O ( (x-x_0)^3).
\ee
Where we can see that when $x^{\mu}=x_0^{\mu}$ we have that $y^{\overline{\mu}}=0$, and hence that our coordinates are centred at $p_0$. The inverse relation to first order is
\be\label{eq:RNCinversetrans}
x^{\mu}=x_0^{\mu}+A^{\mu}_{\;\overline{\nu}}y^{\overline{\nu}}+O (y^3)
\ee
where the $O(y^2)$ term vanishes due to the fact that $\Gamma^{\overline{\mu}}_{\;\overline{\nu}\overline{\rho}} (p_0)=0$. One finds that the $A^{\mu}_{\;\overline{\mu}}$ must satisfy
\be\label{eq:RNCeqnforA}
A^{\overline{\mu}}_{\;\mu}A^{\mu}_{\;\overline{\nu}}=\delta^{\overline{\mu}}_{\;\overline{\nu}}\;,\;\;\;\;A^{\mu}_{\;\overline{\mu}}A^{\overline{\mu}}_{\;\nu}=\delta^{\mu}_{\;\nu}
\ee
Let us denote the time-coordinate of $p$ in GNCs by $T=x^0_p=t_p$. For the transformation from GNCs to RNCs at $p_0$ one finds that $A^{\overline 0}_{\;0}=1$, $A^{\overline 0}_{\;i}=0$ and $\delta_{\ibar\jbar}=A^i_{\;\ibar}A^j_{\;\jbar}h_{ij} (p_0)$, which in particular implies $y^{\overline{0}}_p=\overline{t}_p=T+O(T^3)$.

@@

This is now a simple volume integral but we need to find the boundaries of $\mathcal{X}_p$ to write down the integration limits. There are two parts to the boundary of the (solid) truncated cone: the \emph{base}, defined as $B(p):=\Sigma\cap J^-(p)$, and the \emph{top} defined as the set corresponding to the lightcone emanating at $p$ with all points below $\Sigma$ removed, written as $T(p):=\partial J^-(p)\cap J^+(\Sigma)$. Now the boundary of $\mathcal{X}_p$ is simply $\partial\mathcal{X}_p=B(p)\cup T(p)$. First we look at the top. Following \cite{Khetrapal_Sumati:Causal_Diamond_Volume} it can be shown that the first curvature correction to the lightcone volume comes in at $O (T^{d+2})$ and so can be ignored for our purposes. This means that the top can be treated as an effectively flat lightcone in RNCs, emanating from the point on the time axis at coordinate $T$, and thus it corresponds to the set of points satisfying $ (y^{\overline{1}})^2+ (y^{\overline{2}})^2+...+ (y^{\overline{d-1}})^2= T-\overline{t}$ where the coordinates  $y^{\overline{i}}$ run from $0$ up to some maximum at the points where the flat cone meets the surface $\Sigma$.

>>>>

Let $(M,g)$ be a sufficiently well-behaved $d$-dimensional Lorentzian manifold. We assume that $(M,g)$ has no boundary, is of finite volume, and admits a \emph{spacelike partitioning} surface, $\Sigma$, defined as a spacelike closed compact surface in $M$, such that the causal past and future sets, $M^\pm=J^\pm (\Sigma)$, satisfy $M^+\cap M^-=\Sigma$.~\!\footnote{We do not require that $M^+\cup M^-=M$.} It then follows that $\partial M^\pm = \Sigma$, and further that no points in $M^+\setminus\Sigma$ precede any points in $M^-\setminus\Sigma$.~\!\footnote{If there did exist a $p\in M^+\setminus\Sigma$ and $q\in M^-\setminus\Sigma$ such that $p\preceq q$, then we would have $(q\in J^-(\Sigma) \text{ and } p\preceq q)\implies p\in J^-(\Sigma)\implies p\in M^-$, which would contradict $M^+\cap M^-=\Sigma$.} The Gibbons-Hawking-York boundary term for $M^\pm$ is then given by
\be\label{eq:GHYBT_in_continuum}
{S}^{(d)}_{GHY}\left[M^\pm\right]= \mp \frac{1}{l_p^{d-2}}\int_{\Sigma} K d\Sigma
\ee
where $l_p$ is the rationalised Planck length $l_p=(8\pi G)^{\frac{1}{d-2}}$ ($G$ denoting the gravitational constant in $d$ dimensions) and where $K$ is the trace of the extrinsic curvature $K_{\mu\nu}=h_{\mu}^\rho h_\nu^\sigma \nabla_\rho n_\sigma$ of $\Sigma$, defined here with future-pointing timelike unit normal covector $n_{\mu}=\partial_\mu S/\sqrt{g^{\mu\nu}\partial_\mu S\partial_\nu S}$. (We will work with a mostly plus convention for the metric, in which this translates into a past-pointing normal vector $n^{\mu}$).

@@

Fix $p\in M$ and denote its coordinate values in GNCs by $x^\mu_p= (t_p,\mathbf x_p)$. It will be convenient to use RNCs centered not at the tip, $p$, of the cone but instead at the point $p_0$ where the unique geodesic through $p$ whose tangent is normal to $\Sigma$ intersects $\Sigma$. In GNCs this simply corresponds to the point $p_0$ with coordinates $x_0^\mu=x^\mu (p_0)= (0,\mathbf x_p)$. RNCs centered at $p_0$ will be given the symbol $y^{\overline{\mu}}$, or in terms of coordinates as $(y^{\overline{0}},\mathbf{y})=(\overline{t},\mathbf{y})$.
%\mb{let's put all facts about $x^\mu$ and $y^{\overline\mu}$ for $p$ and $p_0$ here and expansion of det in RNC.} 
We need to assume that for any $p$ in $U_\varepsilon$, the Riemann normal neighbourhood $U_p\subset M$ (throughout which the RNC system centered at $p_0$ is well-defined) contains the truncated cone $J^- (p)\cap J^+ (\Sigma)$. This assumption seems reasonable given the that $U_\varepsilon$ can be made arbitrarily ``thin'' as $\rho\rightarrow\infty$. In RNCs the metric and the Christoffel symbols at $p_0$ are those of flat space:
\be\label{eq:RNCMetricTransAtPAndChris}
g_{\overline{\mu} \overline{\nu}} (p_0)=\eta_{\overline{\mu} \overline{\nu}}=A^{\mu}_{\;\overline{\mu}}A^{\nu}_{\;\overline{\nu}}g_{\mu\nu} (p_0)\;,\;\;\;\;\Gamma^{\overline{\mu}}_{\;\overline{\nu}\overline{\rho}} (p_0)=0
\ee
The $A^{\mu}_{\;\overline{\mu}}$ govern the coordinate transformation from GNCs to RNCs to linear order. %but $O (x^2)$ corrections may be required. 
To second order the coordinate transformation is given by
\be\label{eq:RNCtotaltrans}
y^{\overline{\mu}}=A^{\overline{\mu}}_{\;\nu}(x^\nu -x_0^{\nu})+\frac{1}{2}A^{\overline{\mu}}_{\;\mu}\Gamma^{\mu}_{\;\nu\rho} (p_0)(x^\nu -x_0^{\nu}) (x^\rho -x_0^{\rho})+O ( (x-x_0)^3).
\ee
Where we can see that when $x^{\mu}=x_0^{\mu}$ we have that $y^{\overline{\mu}}=0$, and hence that our coordinates are centred at $p_0$. The inverse coordinate transformation to first order is
\be\label{eq:RNCinversetrans}
x^{\mu}=x_0^{\mu}+A^{\mu}_{\;\overline{\nu}}y^{\overline{\nu}}+O (y^3)
\ee
where the $O(y^2)$ term vanishes due to the fact that $\Gamma^{\overline{\mu}}_{\;\overline{\nu}\overline{\rho}} (p_0)=0$. One finds that the $A^{\mu}_{\;\overline{\mu}}$ must satisfy
\be\label{eq:RNCeqnforA}
A^{\overline{\mu}}_{\;\mu}A^{\mu}_{\;\overline{\nu}}=\delta^{\overline{\mu}}_{\;\overline{\nu}}\;,\;\;\;\;A^{\mu}_{\;\overline{\mu}}A^{\overline{\mu}}_{\;\nu}=\delta^{\mu}_{\;\nu}
\ee
Let us denote the time-coordinate of $p$ in GNCs by $T=x^0_p=t_p$. For the transformation from GNCs to RNCs at $p_0$ one finds that $A^{\overline 0}_{\;0}=1$, $A^{\overline 0}_{\;i}=0$ and $\delta_{\ibar\jbar}=A^i_{\;\ibar}A^j_{\;\jbar}h_{ij} (p_0)$, which in particular implies $y^{\overline{0}}_p=\overline{t}_p=T+O(T^3)$.

@@

This is now a simple volume integral but we need to find the boundaries of $\mathcal{X}_p$ to write down the integration limits. There are two parts to the boundary of the (solid) truncated cone: the \emph{base}, defined as $B(p):=\Sigma\cap J^-(p)$, and the \emph{top} defined as the set corresponding to the lightcone emanating at $p$ with all points below $\Sigma$ removed, written as $T(p):=\partial J^-(p)\cap J^+(\Sigma)$. Now the boundary of $\mathcal{X}_p$ is simply $\partial\mathcal{X}_p=B(p)\cup T(p)$. First we look at the top. Following \cite{Khetrapal_Sumati:Causal_Diamond_Volume} it can be shown that the first curvature correction to the lightcone volume comes in at $O (T^{d+2})$ and so can be ignored for our purposes. This means that the top can be treated effectively as a flat lightcone in RNCs, emanating from the point on the time axis at coordinate $T$, and thus it corresponds to the set of points satisfying $ (y^{\overline{1}})^2+ (y^{\overline{2}})^2+...+ (y^{\overline{d-1}})^2= T-\overline{t}$ where the coordinates  $y^{\overline{i}}$ run from $0$ up to some maximum at the points where the flat cone meets the surface $\Sigma$.
